---
title: "Project Report: Data Sci Doggos"
author: "Naina Ahuja, Mary Knoop, Sam Loewen, Uma Palaniappan, Britney Mazzetta"
date: "December 5, 2019"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
---

```{r setup, message = FALSE, results = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(viridis)
library(leaflet)
```

$~$

# Motivation

The motivation of this project was to compare dog based data across the various boroughs of NYC to deepen our understanding of the current patterns and trends among canines in NYC. We were able to do this by using the NYC DOMHM Dog Licensing Dataset from 2017 - 2019 to accomplish this goal. 

$~$

## Goals of our project

* Understanding and visualizing dog distribution by borough in NYC 
* Identifying trends of licenses by year and/or month in NYC
* Predicting dogs age based on borough and gender of dog 


$~$

# Related Work

We were inspired to work on this project given our collective love for dogs. We were further intrigued to understand the current distribution and trends of dogs within NYC. Drawing inspiration from the NYC Air BnB dataset modeled in class, we realized we could perform similar analysis with this database to better understand what the current licensed dog landscape looks like. We thought that this website would be a knowledge base for any individual who would like to learn more about the current landscape of dogs in NYC.  

$~$

# Initial Questions

We sought to understand the current landscape of licensed dogs in New York City. Additionally, we were also interested in comparing these demographic factors among the five boroughs. Furthermore, we wanted to deepen our understanding on the current patterns of licensing between 2017 - 2019 and examine if there were variations by year or month of license. Finally, we sought to find ways of predicting a dog’s age based on the borough and the gender of the dog. Based on this information, our questions are:  


* What are the risk factors for heart disease and stroke? How can we visualize the correlation of individual risk factors with heart disease and stroke?
* What is the distribution of these risk factors and outcomes in New York city?
* How can we predict and visualize an individual’s risk for heart disease?


$~$

# Data

We utilized the NYC Dog Licensing dataset. All dog owners who reside in NYC are required by law to license their dogs. This is a publicly available dataset sourced from the Department of Health and Mental Hygiene (DOHMH) Dog Licensing System where owners are able to apply for and renew dog licenses. This dataset was created on September 19, 2017 and was last updated on July 25, 2019. There are 10 variables in this dataset with a raw number of 345,727 observations. The variables provided in the raw dataset are as follows: 

* RowNumber: Row Number 
* AnimalName: User-provided dog name (unless specified otherwise) 
* AnimalGender: M(Male) or F(Female) dog gender 
* AnimalBirthMonth: Month and year dog was born 
* BreedName: Dog Breed 
* Borough: Owner Borough 
* ZipCode: Owner Zip Code 
* LiscenceIssuedDate: Date the dog license was issued 
* LiscenceExpiredDate: Date the dog license expires 
* Extract Year: Year Data was extracted

$~$

### Loading and Cleaning Data


$~$

# Exploratory Analysis

$~$

### Loading and tidying the data

The dataset was cleaned and adjusted for this analysis. All dogs whose breed were listed as "unknown" were removed, as well as any dogs who were listed as over 20 years old at the time of thier license. Originally there were no boroughs listed in this dataset, so we used the available zip codes to assign boroughs to each record. Records whose zip code did not correlate to a borough were dropped. A variable `license_length` was added to calculate the lenght in years of each license. There were slight differences in the way breed names were recorded, these were adjusted to be uniform.

Variables were changed to appropriate data type where neccessary.

$~$

```{r load, cache = TRUE, warning = FALSE, message = FALSE}
lat_lon = read.csv("./data/lat-long.csv")

dogz = read.csv("./data/NYC_Dog_Licensing_Dataset.csv") %>% 
  janitor::clean_names() %>% 
  filter(breed_name != "Unknown",
         extract_year-animal_birth_month < 20) %>% 
  mutate(animal_name = str_to_sentence(animal_name),
         borough = ifelse(zip_code %in% c(10001:10282),"Manhattan", borough),
         borough = ifelse(zip_code %in% c(10301:10314), "Staten Island", borough),
         borough = ifelse(zip_code %in% c(10451:10475), "Bronx", borough),
         borough = ifelse(zip_code %in% c(11004,11101:11106,11109,11351,11354:11375,11377:11379,11385,11411:11423,11426:11430,11432:11436,11691:11694,11697), "Queens", borough),
         borough = ifelse(zip_code %in% c(11201,11203:11226,11228:11239,11241:11243,11249,11252,11256), "Brooklyn", borough),
         borough = as.factor(borough),
         breed_name = as.character(breed_name),
         license_issued_date = as.Date(license_issued_date, format = "%m/%d/%Y"),
         license_expired_date = as.Date(license_expired_date, format = "%m/%d/%Y"),
         license_length = year(license_expired_date) - year(license_issued_date),
         dog_age = 2019 - animal_birth_month,
         breed_name = ifelse(breed_name == "American Pit Bull Terrier/Pit Bull", "American Pit Bull Mix / Pit Bull Mix", breed_name),
         breed_name = ifelse(breed_name == "American Pit Bull Terrier/Pit Bull", "American Pit Bull Mix / Pit Bull Mix", breed_name),
         breed_name = ifelse(breed_name == "Australian Cattle Dog", "Australian Cattledog", breed_name),
         breed_name = ifelse(breed_name == "Bassett Hound", "Basset Hound", breed_name),
         breed_name = ifelse(breed_name == "Brittany", "Brittany Spaniel", breed_name),
         breed_name = ifelse(breed_name == "Coonhound, Black and Tan", "Black and Tan Coonhound", breed_name),
         breed_name = ifelse(breed_name == "Coonhound, Black and Tan", "Black and Tan Coonhound", breed_name),
         breed_name = ifelse(breed_name == "Coonhound, Blue Tick", "Bluetick Coonhound", breed_name),
         breed_name = ifelse(breed_name == "Coonhound, Treeing Walker", "Treeing Walker Coonhound", breed_name),
         breed_name = ifelse(breed_name == "Cotton de Tulear", "Coton de Tulear", breed_name),
         breed_name = ifelse(breed_name == "Japanese Chin/Spaniel", "Japanese Chin", breed_name),
         breed_name = ifelse(breed_name == "Jindo Dog, Korea", "Jindo", breed_name),
         breed_name = ifelse(breed_name == "Mastiff, French (Dogue de Bordeaux)", "Dogue de Bordeaux", breed_name),
         breed_name = ifelse(breed_name == "Pharoh hound", "Pharaoh Hound", breed_name),
         breed_name = ifelse(breed_name == "Schipperke", "Schipperkee", breed_name),
         breed_name = ifelse(breed_name == "Schnauzer, Miniature", "Miniature Schnauzer", breed_name),
         breed_name = ifelse(breed_name == "Schnauzer, Miniature Crossbreed", "Miniature Schnauzer, Crossbreed", breed_name),
         breed_name = ifelse(breed_name == "Welsh Corgi, Cardigan", "Cardigan Welsh Corgi", breed_name),
         breed_name = ifelse(breed_name == "Welsh Corgi, Pembroke", "Pembroke Welsh Corgi", breed_name),
         breed_name = ifelse(breed_name == "Bull Dog, French", "French Bulldog", breed_name),
         breed_name = ifelse(breed_name == "Collie, Bearded", "Bearded Collie", breed_name),
         breed_name = ifelse(breed_name == "Bullmastiff", "Mastiff, Bull", breed_name),
         breed_name = ifelse(breed_name == "Neapolitan Mastiff", "Mastiff, Neapolitan", breed_name),
         breed_name = ifelse(breed_name == "Tibetan Mastiff", "Mastiff, Tibetan", breed_name),
         breed_name = ifelse(breed_name == "Pointer, German Shorthaired", "German Shorthaired Pointer", breed_name),
         breed_name = ifelse(breed_name == "Pointer, German Wirehaired", "German Wirehaired Pointer", breed_name),
         breed_name = ifelse(breed_name == "Schnauzer, Giant", "Giant Schnauzer", breed_name),
         breed_name = ifelse(breed_name == "Schnauzer, Standard", "Standard Schnauzer", breed_name),
         breed_name = ifelse(breed_name == "Chinese Shar-Pei", "Shar-Pei, Chinese", breed_name)) %>% 
  drop_na(borough) %>% 
  rename(animal_birth_year = animal_birth_month, license_id = row_number) %>% 
  select(-extract_year)
```

The main database included the same dog multiple times if she had more than one license. We created a subset to catalog only the distinct dogs present in our dataset. To identify individual dogs we matched records on animal name, gender, birth year, and breed name. We also added latitude and longitude data that corresponds to zip code for mapping purposes.

```{r}
dogz_1 =
  dogz %>% 
  distinct(animal_name, animal_gender, animal_birth_year, breed_name, .keep_all = TRUE) %>% 
  select (-license_issued_date, -license_expired_date, -license_length)

distinct_dogz =
  left_join(dogz_1, lat_lon, by = "zip_code") %>% 
  select(-City,-State,-Timezone,-Daylight.savings.time.flag,-geopoint) %>% 
  janitor::clean_names()
```

### Gender of dogs by borough

```{r, warning = FALSE}
plot_2=
  distinct_dogz %>% 
  filter(animal_gender == "M" | animal_gender == "F") %>% 
  ggplot(aes(x = borough, fill = animal_gender)) +
  geom_histogram(position = "dodge", stat = "count") +
  labs(
    title = "Gender of dogs by borough",
    x = "New York Borough",
    y = "Dog Gender"
  )

plot_2
```


### Age Distribution of Dogs by Boroughs 

```{r}
plot_3=
  ggplot(distinct_dogz, aes(x = borough, y = dog_age)) +
  geom_boxplot (color = "purple") +
  labs(
    title = "Age distribution of dogs by borough",
    x = "New York Borough",
    y = "Dog's Age"
  )
plot_3
```

### Top Breed of dogs by Borough 

```{r, message = FALSE}
plot_4=
  distinct_dogz %>% 
  group_by(borough) %>% 
  count(breed_name) %>% 
  top_n(5) %>% 
  ggplot(aes(x = breed_name, y = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_grid(~borough) +
  labs(
    title = "Top 10 breeds of dogs by borough",
    x = "Dog Breed",
    y = "Total Count"
  )
plot_4
```


### Number of licenses over time (license issue date vs. number of licenses) 

```{r}
plot_5.5=
  dogz %>% 
  mutate(license_issued_date = floor_date(as_date(license_issued_date), "month")) %>% 
  group_by(license_issued_date) %>% 
  summarize(n_obs = n()) %>% 
      ggplot(aes(x = license_issued_date, y = n_obs, color = n_obs)) + 
      geom_point() + geom_line() +
      labs(title = "Number of licenses over time", 
      x = "License issue date", 
      y = "Number of licenses")
plot_5.5
```


### Number of licenses issued per month (liscence issue date vs. number of liscences) 

```{r}
plot_5.51=
  dogz %>% 
  mutate(license_issued_date = month(license_issued_date),
         license_issued_date = month.abb[license_issued_date],
         license_issued_date = factor(license_issued_date, c("Jan","Feb", "Mar", "Apr", "May", "Jun", "Jul","Aug", "Sep", "Oct","Nov", "Dec"), ordered = TRUE)) %>%
  group_by(license_issued_date) %>% 
  summarize(n_obs = n()) %>% 
      ggplot(aes(x = license_issued_date, y = n_obs)) + 
      geom_point() + geom_line(group=1) +
      labs(title = "Number of licenses issued per month", 
      x = "License issue date", 
      y = "Number of licenses")
plot_5.51
```


### Most popular dog name by birth year 

```{r, message = FALSE}
top_dogz =
distinct_dogz %>% 
  filter(animal_name != "Unknown",
         animal_name != "Name not provided") %>% 
  group_by(animal_name) %>% 
  summarize(n_obs = n()) %>% 
  arrange(desc(n_obs)) %>% 
  top_n(10)

knitr::kable(top_dogz)

plot_6 = 
distinct_dogz %>% 
  filter(animal_name != "Unknown",
         animal_name != "Name not provided",
         animal_name != "Name",
         animal_birth_year != 1997) %>% 
  group_by(animal_birth_year, animal_name) %>% 
  summarize(n_obs = n()) %>% 
  top_n(1) %>% 
    ggplot(aes(x=animal_birth_year, y=n_obs, group = animal_name, color = animal_name)) + geom_point() + geom_line() +
  labs(title = "Top name by birth year", 
    x = "Birth year", 
    y = "Number of dogs")
plot_6

```

### Map

$~$

# Findings

$~$

# Additional Analysis 

```{r, message = FALSE, warning = FALSE}

#create reference groups that are coded as 
distinct_dogz_reg = 
  distinct_dogz %>% 
  mutate(
    borough = fct_infreq(borough),
    animal_gender = fct_infreq(animal_gender),
    breed_name = fct_infreq(breed_name)
    )
#run linear regression with dog age as Y and borough, breed name and animal gender as categorical
fit = lm(dog_age ~ borough + animal_gender, data = distinct_dogz)

#view outputs
fit %>% 
  broom::tidy()

#create tidy table
fit %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  knitr::kable(digits = 3)
```

```{r}
#graph residuals for each borough
distinct_dogz_reg %>% 
  modelr::add_residuals(fit) %>% 
  ggplot(aes(x = borough, y = resid)) + geom_violin(fill = "pink")
```

$~$

# Discussion